---
title: "Correlation_2 Full Screens"
author: "Robert Reid, modified by Michael Kimble"
date: "4/28/2023"
output: html_document
---

###This code is adapted from Robert Reid "CAS9-volcano.Rmd"; Code is the same until generating 'compared_to_controls'

```{r setup, include=FALSE}
library(tidyverse)
library(ggrepel)
knitr::opts_chunk$set(echo = TRUE)
theme_set(theme_bw())
```

#### 1. Read data

__The data locations__

```{r}
data_loc_list = c(
  "data/2021-10-11-set",
  "data/2021-11-08-set"
)


data = map_dfr(data_loc_list,screenmill::read_screenmill)
```

#### 2. Set analysis parameters

```{r}
technical_replicate_sd_cutoff = 0.3      # Maximum allowable SD for technical replicates
dead_strain_cutoff            = 0.1      # Minimum normalized size of control strain
dead_strain_buffer            = 0.0001   # Amount added to avoid infinite fold change 
slow_growth_cutoff            = 0.3
scale_floor                   = 20
#control                       = "pRS415"
control                       = "g-scr"   #Scrambled gRNA used as control
```

#### 3. plate normalize

```{r echo=FALSE, message=FALSE,warning=FALSE}
plate_normalized <-
  data %>%
  group_by(plate_id) %>%
  mutate(
    size_plate_norm = size / mean(size[plate_control], trim = 0.2)
  ) %>%
  ungroup() %>%
  select (-control_query_id) %>%
  mutate(control_query_id = control)
```

#### 4. stack normalize

```{r echo=FALSE, message=FALSE,warning=FALSE,fig.width=6,fig.height=4}
plate_stack <- 
  plate_normalized %>%
  group_by(colony_row,colony_col) %>%
  summarise(
    stack_size = median(size_plate_norm)
  ) %>%
  ungroup()

stack_model <-
  plate_stack %>%
  filter(stack_size > 0.25) %>% # leaving blanks may work better
  loess(stack_size ~ colony_col + colony_row, data = .,
        degree = 2,normalize = F,span=0.15) # span is the window of data considered 
#                                               for each segment of smoothing

stack_predict <- predict(stack_model,plate_stack)

plate_predict <- cbind(plate_stack,predict = stack_predict)

position_normal <- plate_normalized %>%
  left_join(plate_predict) %>%
  mutate(
    size_pos_norm = size_plate_norm/predict
  )

#write_csv(position_normal, "data/normalized.csv")
```

#### 5. remove locations with missing data

some strain/query combinations with too few ( < 2) for a t test

Count occurrences

```{r}
o_count = position_normal %>%
  group_by(query_id,strain_id) %>%
  count() %>%
  filter(n < 2)

q_count = o_count %>%
  filter(query_id != control)

c_count = o_count %>%
  filter(query_id == control)

clean_normal = position_normal

if (length(q_count[[1]]) > 0) clean_normal = filter(clean_normal,
                                                    !query_id == q_count$query_id |
                                                    !strain_id == q_count$strain_id)

if (length(c_count[[1]]) > 0) clean_normal = filter(clean_normal,
                                                    strain_id != c_count$strain_id)

#write_csv(clean_for_t, "data/clean-normalized.csv")
```

#### 6. Make comparisons

```{r}
#clean_normal = read_csv("data/normalized.csv")

controls <- filter(clean_normal, query_id == control_query_id)
queries  <- filter(clean_normal, query_id %in% c("g1","g6"))
compared_to_controls <-
  inner_join(
    queries %>%
      select(
        date,temperature,strain_collection_id, strain_name, strain_id, plate, row, column,colony_row,colony_col,
        query_id, query_name,size_pos_norm
      ),
    controls %>% 
      select(
        date,temperature,strain_collection_id, strain_name, strain_id, plate, row, column,colony_row,colony_col,
        size_pos_norm
      ),
    by = c("date",
      "temperature","strain_collection_id", "strain_name", "strain_id", "plate", "row", "column","colony_row","colony_col"
    )
  ) %>%
  mutate(
    growth_dif = size_pos_norm.x - size_pos_norm.y,
    growth_ratio = (size_pos_norm.x + dead_strain_buffer)/(size_pos_norm.y + dead_strain_buffer),
    lgr = log2(growth_ratio)
  )

```


### 7. Use 'compared_to_controls' to calculate statistics of each repeat separately by adding "date" to group_by.

```{r}
t_stat <- compared_to_controls %>%
  group_by(date, temperature, strain_name, strain_id, plate, row, column, query_id) %>%
  summarise(
    t        = t.test(size_pos_norm.x,size_pos_norm.y)$statistic,
    p        = t.test(size_pos_norm.x,size_pos_norm.y)$p.value,
    mean_dif = mean(growth_dif,na.rm=T),
    sd_dif   = sd(growth_dif,na.rm=T),
    mean_lgr = mean(lgr,na.rm=T),
    sd_lgr   = sd(lgr,na.rm=T)
  ) %>%
  filter(!strain_name %in% c("blank","his3"))
```


### 8. Make a data frame containing mean_dif values for the two trials

```{r}
mean_dif_data <- t_stat %>%
  group_by(strain_name, strain_id, plate, row, column, query_id) %>%
  summarize(mean_dif_Trial1 = mean(mean_dif[date == "2021-10-11"]),
            mean_dif_Trial2 = mean(mean_dif[date == "2021-11-08"]))

#write_csv(mean_dif_data, "data/mean_dif_by_trial.csv")
```


### 9. Plot mean_dif values for the two trials. 2021-10-11 is on the x axis and 2021-11-08 is on the y axis. g1 and g6 are plotted separately.

```{r}
ggplot(mean_dif_data, aes(x = mean_dif_Trial1, y = mean_dif_Trial2, color=query_id)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "black") +
  labs(x = "Growth Difference R1", y = "Growth Difference R2", title = "Growth Difference Correlation Analysis") +
  theme_minimal() +
  scale_color_viridis_d(begin=0.2,end = 0.9,option = "turbo") +
  facet_wrap(~query_id) +
  xlim(c(-1.5,1.5)) +
  theme(panel.border = element_rect(color = "gray", fill = NA, size = 1), 
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        strip.text = element_text(size = 10))
```

### 10. Same plot, but with a best fit line using geom_smooth. Slopes and R-squared are noted at the bottom of this chunk.
```{r}
ggplot(mean_dif_data, aes(x = mean_dif_Trial1, y = mean_dif_Trial2, color=query_id)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black", aes(group = query_id)) +
  labs(x = "Growth Difference R1", y = "Growth Difference R2", title = "Growth Difference Correlation Analysis") +
  theme_minimal() +
  scale_color_viridis_d(begin=0.2,end = 0.9,option = "turbo") +
  facet_wrap(~query_id) +
  xlim(c(-1.5,1.5)) +
  theme(panel.border = element_rect(color = "gray", fill = NA, size = 1), 
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        strip.text = element_text(size = 10))

lapply(split(mean_dif_data, mean_dif_data$query_id), function(df) 
  {coef(summary(lm(mean_dif_Trial2 ~ mean_dif_Trial1, df)))
  })

lapply(split(mean_dif_data, mean_dif_data$query_id), function(df) {
  model <- lm(mean_dif_Trial2 ~ mean_dif_Trial1, df)
  summary(model)$r.squared
})


#g1:Slope = 0.162 and R-squared = 0.032
#g6:Slope = 0.235 and R-squared = 0.060
```





### The same correlation analysis can be done for mean log ratio
### Make a data frame containing mean_lgr values for the two trials

```{r}
mean_lgr_data <- t_stat %>%
  group_by(strain_name, strain_id, plate, row, column, query_id) %>%
  summarize(mean_lgr_Trial1 = mean(mean_lgr[date == "2021-10-11"]),
            mean_lgr_Trial2 = mean(mean_lgr[date == "2021-11-08"]))
```

### Plot mean_lgr values for the two trials. 2021-10-11 is on the x axis and 2021-11-08 is on the y axis. g1 and g6 are plotted separately.

```{r}
ggplot(mean_lgr_data, aes(x = mean_lgr_Trial1, y = mean_lgr_Trial2, color=query_id)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "black") +
  labs(x = "Log Growth Ratio R1", y = "Log Growth Ratio R2", title = "Log Growth Ratio Correlation Analysis") +
  theme_minimal() +
  scale_color_viridis_d(begin=0.2,end = 0.9,option = "turbo") +
  facet_wrap(~query_id) +
  xlim(c(-7,7)) +
  theme(panel.border = element_rect(color = "gray", fill = NA, size = 1), 
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        strip.text = element_text(size = 10)) 
```


### Same plot, but with a best fit line using geom_smooth. Slopes and R-squared are noted at the bottom of this chunk.
```{r}
ggplot(mean_lgr_data, aes(x = mean_lgr_Trial1, y = mean_lgr_Trial2, color=query_id)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black", aes(group = query_id)) +
  labs(x = "Log Growth Ratio R1", y = "Log Growth Ratio R2", title = "Log Growth Ratio Correlation Analysis") +
  theme_minimal() +
  scale_color_viridis_d(begin=0.2,end = 0.9,option = "turbo") +
  facet_wrap(~query_id) +
  xlim(c(-7,7)) +
  theme(panel.border = element_rect(color = "gray", fill = NA, size = 1), 
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        strip.text = element_text(size = 10))

lapply(split(mean_lgr_data, mean_lgr_data$query_id), function(df) 
  {coef(summary(lm(mean_lgr_Trial2 ~ mean_lgr_Trial1, df)))
  })

lapply(split(mean_lgr_data, mean_lgr_data$query_id), function(df) {
  model <- lm(mean_lgr_Trial2 ~ mean_lgr_Trial1, df)
  summary(model)$r.squared
})


#g1:Slope = 0.143 and R-squared = 0.033
#g6:Slope = 0.289 and R-squared = 0.108
```

##Matching color to volcano plots

```{r}
ggplot(mean_dif_data, aes(x = mean_dif_Trial1, y = mean_dif_Trial2)) +
  geom_point(aes(color = query_id)) +
  scale_color_manual(values = c("red", "blue")) +
  geom_abline(intercept = 0, slope = 1, color = "black") +
  labs(x = "Growth Difference R1", y = "Growth Difference R2", title = "Growth Difference Correlation Analysis") +
  theme_bw() +
  facet_wrap(~query_id) +
  xlim(c(-1.5,1.5)) +
  theme(panel.border = element_rect(color = "gray", fill = NA, size = 1), 
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        strip.text = element_text(size = 10))
```


